---
title: 'chapter 2 Getting Started with Data in R'
output:
  tufte::tufte_html:
    tufte_features: ['fonts','background']
    toc: true
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = FALSE)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(extraDistr)
library(gridExtra)
library(latex2exp)
library(moments)
library(bookdown)
library(rsconnect)
library(tufte)
```

\newcommand{\E}{\text{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\SD}{\text{SD}}
\newcommand{\SE}{\text{SE}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\Cor}{\text{Cor}}
\renewcommand{\P}{\text{P}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\sumin}{\sum_i^n}
\newcommand{\Bias}{\text{Bias}}

--- 

# Loading Data

There are many functions in R for loading data of various formats: 

* `read.csv()` for csv files (comma separated values) 
* `read.tsv()` for tsv files (tab separated values) 
* `read.xlsx()` for Excel files (from the `readxl` package)
* `read.dta13()` for dat files (from the `readstata13` package)
* `read.table()` for huge datasets

When loading data you must specify the exact file path in the argument (see below). If you don't know how to find your file path, give it a google. Remember to name your dataset (i.e. assign your dataset to an object). 

The following code loads a csv file with data on the body temperature, heart rate, and gender of 130 subjects.^[Download the data <a href="base64 data" download="./data/normtemp.csv">here</a>. Excuse its mediocrity--there will (hopefully) be more interesting examples in chapters to come. To see where the data actually comes from, click <a href="https://www.ncbi.nlm.nih.gov/pubmed/1302471">here.</a>] Locally the file is called 'normtemp.csv' and it is assigned to an object called `cardiacdata`:  

```{r}
cardiacdata <- read.csv('./data/normtemp.csv', header = TRUE)
```

Note the extra argument, `header = TRUE`, which specifies that the first row of the dataset is a header. If your dataset has no header you should specify `header = FALSE`.^[Here the `=` operator is not used for variable assignment, but rather to specify a parameter for the `read.csv()` function (this is the main difference between the `<-` and `=` operators).]  



\ 

---

# Viewing Data 

To view the entire dataset, use the `View()` command in the console. A table view of the dataset will open as a new tab. Don't use the `View()` command for large datasets as it is very memory intensive.  

Another way to see your data is to print the first or last few rows using the `head()` or `tail()` function. You can specify exactly how many rows as an additional argument (by default it will print six):  

```{r}
head(cardiacdata, n = 5)
```

To check the column names of your dataset, use `colnames()`:

```{r}
colnames(cardiacdata)
```

To check the dimensions of your dataset (number of rows and columns), use `dim()`:

```{r}
dim(cardiacdata)
```



\ 

--- 

# Basic Data Structures in R

In statistics a **variable** usually refers to something that is measured. Data is just observations of some variable--in the cardiac dataset each column contains data pertaining to a specific variable. Data can be continuous, discrete, or categorical.   

- **Continuous data** can take on infinitely many values (real numbers). E.g. the `bodytemp` variable in the cardiac dataset. 
- **Discrete data** can take on countable values only (integer numbers). E.g. the `heartrate` variable in the cardiac dataset. 
- **Categorical data** fall into a finite number of categories or distinct groups. E.g. the `gender` variable in the cardiac dataset.^[Even though `gender` is numerically coded in the temperatures dataset, it is still categorical since it only takes on two values, 1 and 2.]  

Everything in R is an object. When you load data into R, it is parsed into objects. Every object has a **data type**, and different forms of data are parsed into different data types.  

Below are the five elementary data types in R: 

* __character__ -- e.g. `'abcd'`
* __integer__ -- integer numbers, e.g. `'2'`
* __numeric__ -- decimal numbers, e.g. `'2.21'` 
* __complex__ -- complex numbers, e.g. `'2+2i'`
* __logical__ -- either `TRUE` or `FALSE`  

Objects may be combined to form larger data structures. Some common ones:  

* __vector__ -- a one-dimensional array; there are two kinds of vectors:
    + __atomic vector__ -- holds data of a single data type
    + __list__ -- holds data of multiple data types
* __matrix__ -- a two-dimensional array; all columns have the same data type
* __data frame__ -- a two-dimensional array; columns may have different data types  

You can check an object's data type using `class()`:

```{r}
class(cardiacdata)
```

The data frame is indeed a common structure for tabular data. To check the data type(s) in the column `bodytemp`:

`r margin_note("Note the extract operator, $\\$$ , which is used to extract a named object from a larger data structure--in this case extract a column from a data frame.")`

```{r}
class(cardiacdata$bodytemp)
```

Similarly for the column `heartrate`:

```{r}
class(cardiacdata$heartrate)
```



\ 

---

\ 


